Explanatory Data Analysis of Trade Statistics of Turkey

### 2018-12-18 Draft/Draft20181218.Rmd
.

---
title: "Group Fou\\*R* Term Project"
subtitle: "MEF University - BDA 503 Essentials of Data Analytics - Fall 2018"
date: "December 2018"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: false
    df_print: paged
    code_folding: hide

---
# Explanatory Data Analysis of Trade Statistics of Turkey {.tabset .tabset-fade .tabset-pills}

@@

We worked on trade statistics data of Turkey against the G7 countries and 3 more countries (Russia, India, China) both for export and import.Our aim is to analyze Turkey's trade characteristics by product groups. Within this aim, we take trade values and quantities by product groups in MS Excel file format from trademap.org website for each trade partner country for the period of 2013 - 2017. 
In this study, We applied data manipulation to make it proper for analysis and explanatory data analysis with visulations and conclude with overall statistical overview of Turkey's trade.

@@

Following the extraction from source (Trademap.org), our raw data consist of 10 countries' trade data against Turkey. This data splits to 2 with trade direction as 'Export' and 'Import'. The other dimension is stat for trade, one is 'Quantity' of Export/Import, the other is 'Value' in thousand USD of Export/Import. Therefore we have four main data frame for each country (total of 40 excel files).

@@

Table structures for "Quantity" and "Value" files are as follows:

@@

Table 1: Quantity excel table structure

@@

Table 2: Values excel table structure

In addition to main raw data, we also have look-up table for product codes; This lookup table registers product labels in 2 digit code that is less categorized than main data.

@@

<tbody>
<tr class="odd">
<td align="left">'01</td>
<td align="left">Live animals</td>
</tr>
<tr class="odd">
<td align="left">'02</td>
<td align="left">Meat and edible meat offal</td>
</tr>
<tr class="odd">
<td align="left">'03</td>
<td align="left">Fish and crustaceans, molluscs and other aquatic invertebrates</td>
</tr>
<tr class="odd">
<td align="left">'04</td>
<td align="left">Dairy produce; birds' eggs; natural honey; edible products of animal origin, not elsewhere ...</td>
</tr>

@@

Table 3: Product look-up excel table

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}

library(shiny)
library(tidyverse)
library(dplyr)
library(readxl)
library(leaflet)
library(reshape2)
library(gridExtra)
library(lubridate)
library(plotrix)
```

In order to read these excel files, firstly we create a function named "getDataSet":

@@

colnames(import_quantity_dataset) <- c("Country","Product_Code","Product_Label","Quantity_2013","Unit1","Quantity_2014","Unit2","Quantity_2015","Unit3","Quantity_2016","Unit4","Quantity_2017","Unit5")
colnames(import_value_dataset) <- c("Country","Product_Code","Product_Label","Value_2013","Value_2014","Value_2015","Value_2016","Value_2017")
colnames(export_quantity_dataset) <- c("Country","Product_Code","Product_Label","Quantity_2013","Unit1","Quantity_2014","Unit2","Quantity_2015","Unit3","Quantity_2016","Unit4","Quantity_2017","Unit5")
colnames(export_value_dataset) <- c("Country","Product_Code","Product_Label","Value_2013","Value_2014","Value_2015","Value_2016","Value_2017")

@@

And by using the getDataSet function, we merged the Turkey's export-quantity, export-value, import-quantity, import-value tables for 10 selected countries.

```{r}
glimpse(export_quantity_dataset)
glimpse(export_value_dataset)
glimpse(import_quantity_dataset)
glimpse(import_value_dataset)
```

We have 13 and 8 variables in quantity and value tables respectively both for export and import.

@@

Also, we need to join product labels in 2 digit product codes to 4 digit product codes and labels.

```{r, warning=FALSE}
#read the ProductCodes excel
file = "C:/Users/emrek/Google Drive/BDA/503-EssentialsOfDataAnalytics/GitHub/gpj18-group_four/2digitproductlist.xlsx"
prdCode <- read_excel(file, col_names=FALSE, skip=1)
prdCode$X__1 <- gsub("'", "", prdCode$X__1)

@@

Now we have joined tables for import and export stats of Turkey.

@@

But there is a row in Product_Code column as "TOTAL", we must clear it.

```{r}
export_jTable <- export_jTable %>% filter(!grepl("TOTAL", Product_Code))
import_jTable <- import_jTable %>% filter(!grepl("TOTAL", Product_Code))
```

We have 16 variables in these tables however quantity and value variables repeat for each year. So actually, it is better to organize the table that it will have variables "Year", "Quantity", "Value" instead of 10 columns.

@@

Now, we have more useful tables, with 9 variables and 61100 rows.

@@

@@

```{r}
ttable1=export_jTableMELTED%>%group_by(Country)%>%
  summarise(co_total=sum(Values,na.rm = TRUE))%>%
  arrange(desc(co_total,year))

@@

pex1_table<-ttable1%>%
  mutate(perc_count=(co_total/sum(ttable1$co_total))*100)%>%
  select(Country,co_total,perc_count)

@@

pim1_table<-imptable1%>%
  mutate(perc_count=(co_total/sum(ttable1$co_total))*100)%>%
  select(Country,co_total,perc_count)

lbls_i <- paste(pim1_table$Country,round(pim1_table$perc_count,1)) # add percents to labels
lbls_i <- paste(lbls_i,"%",sep="") # ad % to labels

pie3D(pim1_table$perc_count,labels=lbls_i,main = "Total Percentage Import Per Countries", explode=0.1, radius=.9,
      labelcex = 1.2, start=0.7)

lbls_e <- paste(pex1_table$Country,round(pex1_table$perc_count,1)) # add percents to labels
lbls_e <- paste(lbls_e,"%",sep="") # ad % to labels

pie3D(pex1_table$perc_count,labels=lbls_e,main = "Total Percentage Export Per Countries", explode=0.1, radius=.9,
             labelcex = 1.2, start=0.7)

At import side, Turkeys' biggest trade partners are China, Germany, Russia and US while they are Germany, UK, Italy and USA at export side. Germany and USA are the 2 countries exist for both type of trade.

When we breakdown this info to years:

```{r}
# Export Values in $ 
ttable=export_jTableMELTED%>%group_by(Country,year)%>%
  summarise(co_total=sum(Values,na.rm = TRUE))%>%
  arrange(desc(co_total,year))
#Import Values in $
vtable=import_jTableMELTED%>%group_by(Country,year)%>%summarise(co_total=sum(Values,na.rm = TRUE))%>%arrange(desc(co_total,year))

@@

Except Russia, fluctuates over years are not seem so significant.

In order to deeply understand the data; we build up shiny app with Country, Product and Year variables to review export and import values.

Please follow the link for interactive graph: [Link](https://groupfour.shinyapps.io/tradeOrg)

#### YORUM VE IMAGE EKLENECEK

### Analysis of Year by Year Change in Trade Values

```{r}
tableExport <- export_jTableMELTED %>% 
  group_by(Country, year) %>%
  summarise(totalExpV=sum(Values,na.rm=TRUE)) %>%
  arrange(Country, year) %>%
  mutate(ExpValChng = (totalExpV - lag(totalExpV))/lag(totalExpV)*100)

@@

```{r, }
ggplot(table, aes(x=year)) +
  geom_point(aes(y=value, color=Country, shape=variable), size=4) + 
  facet_grid(~variable) +
   labs(title="YoY Total Value Change",
    x="Year",
    y="% Change in  Value",
    shape=c("Trade Type"))
```

```{r}

@@

```{r, echo=FALSE}
# Import Analysis

@@

```{r, echo=FALSE}

@@

```{r, echo=FALSE, fig.height=7, fig.width=10 }

@@

@@

To discovery that country's effects to trade deficit, we calculated net_value which formulated as 'Export_value - Import_value'. And visualized that value on the basis of year.

First, group data;

@@

Continue with visulation;

```{r}
####Net Value By Country###
Net_Value_By_Country <- Export_Import_JTable %>%
group_by(Country,year)%>%
summarize (Export_Value = sum(Export_Value), Import_Value = sum(Import_Value), 
           Net_Value = sum(Export_Value - Import_Value),Net_value_K = round((sum(Export_Value - Import_Value)/1000),digits=0))

@@

As seen in the graph, only UK gives trade surplus for all years analysed and China, Russia, Germany and USA have the most significant effect on trade deficit.

@@

The effects of countries are changing year by year. So that as an alternative perspective we could average net value for all countries in five years and normalized that value to show which country's effects are more than others.

```{r }
#Normalized of Average Values for five years
Normalized_Values_By_Country <- Export_Import_JTable %>%
  #filter (year == 2017)%>%
  group_by(Country)%>%
  summarize (Net_Value = mean(Export_Value - Import_Value))%>%
  mutate(normalized_Net_value =  round((Net_Value  - mean(Net_Value))/sd(Net_Value ),2)) %>%
  mutate(value_Net_type =  ifelse(normalized_Net_value < 0, "below", "above"))

@@

Please follow the link for interactive graph: [Link](https://groupfour.shinyapps.io/tradeOrgProAnly/)

@@

#### IMAGE EKLENECEK

@@

Read the USD and BIST100 index values and rename the columns

```{r}
fileUSD= "C:/Users/emrek/Google Drive/BDA/503-EssentialsOfDataAnalytics/GitHub/gpj18-group_four/EVDS.xlsx"
usdR <- read_excel(fileUSD, col_names=FALSE, skip=1)

@@

Draw the graph to represent USD/TL rate, BIST100 index and Trade Deficit in time series.

```{r}
sumImport<- import_jTable %>% group_by(Country) %>% summarise(Value_2013=sum(Value_2013),Value_2014=sum(Value_2014),Value_2015=sum(Value_2015),Value_2016=sum(Value_2016),Value_2017=sum(Value_2017))
sumExport<- export_jTable %>% group_by(Country) %>% summarise(Value_2013=sum(Value_2013),Value_2014=sum(Value_2014),Value_2015=sum(Value_2015),Value_2016=sum(Value_2016),Value_2017=sum(Value_2017))
joinedTableDeficit<-sumExport %>%  inner_join(sumImport,by="Country") %>%
  mutate(Value_2013=Value_2013.y-Value_2013.x,Value_2014=Value_2014.y-Value_2014.x, Value_2015=Value_2015.y-Value_2015.x, Value_2016=Value_2016.y-Value_2016.x,
         Value_2017=Value_2017.y-Value_2017.x) %>%
  select(Country, Value_2013, Value_2014, Value_2015, Value_2016, Value_2017)

@@

Trade deficit has negatively correlated with USD/TL rate between 2013 and 2016 as expected and for the year 2017 we saw the effect of stagflation - while the USD/TL rate increasing (accordingly we know that inflation rate is strongly positively correlatied with the inflation rate (inf rate: 2015-12:5.71%, 2016-12:9.94%, 2017-12:15.47%)) and weak or negative economic growth. 
Besides, there is not a direct correlation between BIST100 index and Trade Deficit Values.

>>>>

---
title: "Explanatory Data Analysis of Trade Statistics of Turkey"
subtitle: "MEF University - BDA 503 Essentials of Data Analytics - Fall 2018 - Group Fou\\*R* Term Project"
date: "December 2018"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: false
    df_print: paged
    code_folding: hide

---
# {.tabset .tabset-fade .tabset-pills}

@@

In this term project, trade statistics data of Turkey against the G7 countries and 3 more countries (Russia, India, China) is analysed. The aim of the project is to analyze Turkey's export and import characteristics by product groups considering the trade value and quantities. Within this aim, we take trade values and quantities by product groups in MS Excel file format from [trademap.org](https://trademap.org/Index.aspx) website for each trade partner for the period between 2013 and 2017. 
In this study, We followed the steps of explanation of raw data, preparation of raw data through merging the raw data files, cleaning, and manipulating,  analysing the data to explore the trade characteristics along with visulations and lastly concluding the comments from analysis.

@@

Following the extraction from source [(Trademap.org)](https://trademap.org/Index.aspx), our raw data consist of 10 countries' trade data against Turkey. This data split with trade direction as 'Export' and 'Import', in other words each country has 2 files as export and import. The other dimension is trade data; one is 'Quantity' of Export/Import, the other is 'Value' in thousand USD of Export/Import, which means each country has one file for 'Quantity' and one file for 'Trade Value'. Therefore we have four main data table for each country (total of 40 excel files).

@@

Variables in "Quantity" and "Value" files are as follows:

@@

@@

In addition to main raw data, we also have a look-up table for product codes. This lookup table registers product labels in 2 digits code that is less categorized than main data.

@@

<tbody>
<tr class="odd">
<td align="left">'01</td>
<td align="left">Live animals</td>
</tr>
<tr class="odd">
<td align="left">'02</td>
<td align="left">Meat and edible meat offal</td>
</tr>

@@

@@

We started with loading the needed libraries for preparation, manipulation, analysis and visualisation.  
```{r, message=FALSE, warning=FALSE}
library(shiny)
library(tidyverse)
library(dplyr)
library(readxl)
library(leaflet)
library(reshape2)
library(gridExtra)
library(lubridate)
library(plotrix)
library(treemap)
```

First of all, we need to read the excel files constituting our raw data. For this aim, we wrote a function named "getDataSet".

@@

This function reads the excel files in the defined path according to given pattern in the filename and bind them in one file named "resultDataset". By using the getDataSet function, we merged the Turkey's export-quantity, export-value, import-quantity and import-value tables for 10 selected countries.

Then, we named the columns and removed the single quotation marks in product code column in each file.

```{r}
colnames(import_quantity_dataset) <- c("Country","Product_Code","Product_Label","Quantity_2013","Unit1","Quantity_2014","Unit2","Quantity_2015","Unit3","Quantity_2016","Unit4","Quantity_2017","Unit5")
colnames(import_value_dataset) <- c("Country","Product_Code","Product_Label","Value_2013","Value_2014","Value_2015","Value_2016","Value_2017")
colnames(export_quantity_dataset) <- c("Country","Product_Code","Product_Label","Quantity_2013","Unit1","Quantity_2014","Unit2","Quantity_2015","Unit3","Quantity_2016","Unit4","Quantity_2017","Unit5")
colnames(export_value_dataset) <- c("Country","Product_Code","Product_Label","Value_2013","Value_2014","Value_2015","Value_2016","Value_2017")

@@

After the first step, the view of export & import quantity and export& import value datasets are below in respective order:

```{r}
glimpse(export_quantity_dataset)
glimpse(import_quantity_dataset)
glimpse(export_value_dataset)
glimpse(import_value_dataset)
```

We have 13 and 8 variables in quantity and value tables respectively for export and import.

@@

Also, we need to join product labels in 2 digit product codes with these table.

```{r, echo=TRUE, warning=FALSE}
#read the ProductCodes excel
file = "C:/Users/emrek/Google Drive/BDA/503-EssentialsOfDataAnalytics/GitHub/gpj18-group_four/2digitproductlist.xlsx"
prdCode <- read_excel(file, col_names=FALSE, skip=1)
prdCode$X__1 <- gsub("'", "", prdCode$X__1)

@@

Now we have joined tables for import and export stats of Turkey. Lets glimpse them:

@@

But there is a row in Product_Code column as "TOTAL", we cleared it as well.

```{r, echo=TRUE}
export_jTable <- export_jTable %>% filter(!grepl("TOTAL", Product_Code))
import_jTable <- import_jTable %>% filter(!grepl("TOTAL", Product_Code))
```

We have 16 variables in these tables however quantity and value variables repeat for each year. So actually, it is better to organize the table that it will have 3 variables as "Year", "Quantity" and "Value" instead of 10 columns.

@@

After above coding, now we have more useful tables, with 9 variables and 61100 rows.

@@

These tables are ready for analysis that will be conducted in the next section. However, before proceeding first lets check the summary of export and import tables.

```{r}
summary(export_jTableMELTED)
sd(export_jTableMELTED$Values)

summary(import_jTableMELTED)
sd(import_jTableMELTED$Values)
```
Since the quantities have different respective units, it is not logical to comment on descriptive statistics for quantity but for the values column we can say that the values are so dispersed with 4391 mean value and  36438 standart deviation for exports and 9495 mean and 110904 standart deviation.

@@

Our analysis starts with to understandthe shares of trade volume of the selected 10 countries on Turkey's import and export.

```{r, fig.align="center"}
# TREEMAP FOR EXPORT

@@

pex1_table<-exptable1%>%
  mutate(perc_count=(co_total/sum(exptable1$co_total))*100)%>%
  select(Country,co_total,perc_count)

treemap(pex1_table,index = c("Country"),vSize = "perc_count",vColor = "perc_count",title="Total Export Percentage on Countries",
        fontsize.title = 17,type="value",palette = "Set1")

# TREEMAP FOR IMPORT

@@

pim1_table<-imptable1%>%
  mutate(perc_count=(co_total/sum(imptable1$co_total))*100)%>%
  select(Country,co_total,perc_count)

treemap(pim1_table,index = c("Country"),vSize = "perc_count",vColor = "perc_count",title="Total Import Percentage on Countries",
        fontsize.title = 17,type="value",palette = "Set1")
```

As seen above, Turkeys' biggest trade partners are China, Germany, Russia and US at import side while they are Germany, UK, Italy and USA at export side. Germany and USA are the 2 top countries seen in both directions of trade.

When we breakdown this info to years, it is seen in below graphs that except Russia, fluctuations over years are not seem so significant but it is worth to deep dive Year over Year (YoY) changes.

```{r, fig.align="center"}
# Export Values in $ 
ttable=export_jTableMELTED%>%group_by(Country,year)%>%
  summarise(co_total=sum(Values,na.rm = TRUE))%>%
  arrange(desc(co_total,year))
#Import Values in $
vtable=import_jTableMELTED%>%group_by(Country,year)%>%summarise(co_total=sum(Values,na.rm = TRUE))%>%arrange(desc(co_total,year))

@@

### Analysis of Year over Year Changes in Trade Values

In order to review YoY changes, we calculated delta of trade value for consecutive years and plotted as below seperating export and import.

```{r, message=FALSE}
tableExport <- export_jTableMELTED %>% 
  group_by(Country, year) %>%
  summarise(totalExpV=sum(Values,na.rm=TRUE)) %>%
  arrange(Country, year) %>%
  mutate(ExpValChng = (totalExpV - lag(totalExpV))/lag(totalExpV)*100)

@@

```{r, fig.align="center"}
ggplot(table, aes(x=year)) +
  geom_point(aes(y=value, color=Country, shape=variable), size=4) + 
  facet_grid(~variable) +
   labs(title="YoY Total Value Change",
    x="Year",
    y="% Change in  Value",
    shape=c("Trade Type"))
```

From the graph, it is observable that especially in the years of 2015, 2016 and 2017 exports from Russia and Canada deviates strongly, up to above %50. And for the import side, again in 2016 and 2017 seems that Russia and Canada are top deviators and suprisingly in 2017 imports from Canada rises more than 100%.

So, we should check which products drives these fluctuations. In order to visualize it, we found the highest delta products through filtering the ones exceeding 2 interquantile range (IQR - the difference between third and first quantile).

```{r, fig.align="center"}

@@

```{r, echo=FALSE, fig.align="center"}
# Import Analysis

@@

```{r, echo=FALSE, fig.align="center"}

@@

```{r, echo=FALSE, fig.height=7, fig.width=10, fig.align="center"}

@@

It is concluded in above graphs that;

* The observations of change of export to Canada in 2015 and 2017 are due to export of "Iron and Steel".
* The observation of change of import from Canada in 2017 is due to import of "Natural or cultured pearls...".
* The decrease in export to Russia in 2015 drives by "Vehicles other than railway..".
* The decrease in export to Russia in 2016 and increase in 2017 is the effect of "Edible fruits and vegetables".
* Imports from Russia deviates due to "Commodities not elsewhere speecified".

Besides this analysis, we can discover more on this topic with different perspectives. To do this, we decided to build up shiny app with dependents of Country, Product and Year. Please follow the link to access to Shiny app.
__[Link](https://groupfour.shinyapps.io/tradeOrg)__

![Edible vegetable export to Russia](C:\Users\emrek\Google Drive\BDA\503-EssentialsOfDataAnalytics\GitHub\gpj18-group_four\images\imageShiny1.png)

This image taken from Shiny app and presents the trade values of edible vegetables over years. It is observable that, in 2016 Turkey's edible vegetable export to Russia dramatically downs.

![Russian sanctions to Turkey](C:\Users\emrek\Google Drive\BDA\503-EssentialsOfDataAnalytics\GitHub\gpj18-group_four\images\Capture.png)

@@

To discover the country's effects to trade deficit, we calculated net_value which formulated as 'Export_value - Import_value'. And visualized that value on the basis of year.

First, we grouped and manipulated the data and continued with visulation;

@@

```{r, fig.align="center"}
####Net Value By Country###
Net_Value_By_Country <- Export_Import_JTable %>%
group_by(Country,year)%>%
summarize (Export_Value = sum(Export_Value), Import_Value = sum(Import_Value), 
           Net_Value = sum(Export_Value - Import_Value),Net_value_K = round((sum(Export_Value - Import_Value)/1000),digits=0))

@@

As seen in the graph, only UK gives trade surplus for all years analysed. On the other hand China, Russia, Germany and USA have the most significant effect on trade deficit.

@@

The effects of countries are changing year by year. So that, as an alternative perspective we decided to take average net value for all countries in five years and normalized that value to show which country's effects are more than others.

```{r , fig.align="center"}
#Normalized of Average Values for five years
Normalized_Values_By_Country <- Export_Import_JTable %>%
  #filter (year == 2017)%>%
  group_by(Country)%>%
  summarize (Net_Value = mean(Export_Value - Import_Value))%>%
  mutate(normalized_Net_value =  round((Net_Value  - mean(Net_Value))/sd(Net_Value ),2)) %>%
  mutate(value_Net_type =  ifelse(normalized_Net_value < 0, "below", "above"))

@@

Please follow the link for interactive graph: __[Link](https://groupfour.shinyapps.io/tradeOrgProAnly/)__

@@

#### PIE GRAFIGI DEGISTIRECEGIZ DİYE SHINY DEN IMAGE LARI ALMADIM HENUZ

@@

First we read the USD and BIST100 index values and rename the columns; Then proceeded with drawing the graph to represent USD/TL rate, BIST100 index and Trade Deficit in time series.

```{r, fig.align="center"}
fileUSD= "C:/Users/emrek/Google Drive/BDA/503-EssentialsOfDataAnalytics/GitHub/gpj18-group_four/EVDS.xlsx"
usdR <- read_excel(fileUSD, col_names=FALSE, skip=1)

@@

```{r, fig.align="center"}
sumImport<- import_jTable %>% group_by(Country) %>% summarise(Value_2013=sum(Value_2013),Value_2014=sum(Value_2014),Value_2015=sum(Value_2015),Value_2016=sum(Value_2016),Value_2017=sum(Value_2017))
sumExport<- export_jTable %>% group_by(Country) %>% summarise(Value_2013=sum(Value_2013),Value_2014=sum(Value_2014),Value_2015=sum(Value_2015),Value_2016=sum(Value_2016),Value_2017=sum(Value_2017))
joinedTableDeficit<-sumExport %>%  inner_join(sumImport,by="Country") %>%
  mutate(Value_2013=Value_2013.y-Value_2013.x,Value_2014=Value_2014.y-Value_2014.x, Value_2015=Value_2015.y-Value_2015.x, Value_2016=Value_2016.y-Value_2016.x,
         Value_2017=Value_2017.y-Value_2017.x) %>%
  select(Country, Value_2013, Value_2014, Value_2015, Value_2016, Value_2017)

@@

It is concluded that trade deficit has negatively correlated with USD/TL rate between 2013 and 2016 as expected and for the year 2017 we saw the effect of stagflation - while the USD/TL rate increasing (accordingly we know that inflation rate is strongly positively correlatied with the inflation rate (inf rate: 2015-12:5.71%, 2016-12:9.94%, 2017-12:15.47%)) and weak or negative economic growth. 
Besides, there is not a direct correlation between BIST100 index and Trade Deficit Values.

## Conclusion

## References
